---
title: "ADA Final exam"
output: html_document
---

# Exercise 1.
In this exercise, we are going to evaluate whether streptomycin improves sore throat using data from a randomized controlled trial that was published in 1948. The original paper can be accessed at https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2091872/?page=1. Please run the code below to get started.
```{r}
install.packages("medicaldata")
install.packages("multcomp")
library(medicaldata)
library(odds.n.ends)
library("MASS")
library("nnet")
library(multcomp)

#assign strep dataset in the medical data package to the environment
strep_tb<-strep_tb #See https://higgi13425.github.io/medicaldata/
head(strep_tb)

#label radnum variable that measures improvement on a six point scale with Death being 1 and Considerable improvement being 6
rad_numLabels <- c("Death","Considerable deterioration","Moderate or slight deterioration","No material change","Moderate or slight improvement","Considerable improvement")

rad_numValues <- 1:6
strep_tb$rad_numLabels <- rad_numLabels[match(strep_tb$rad_num,rad_numValues)]
strep_tb$rad_numLabels <- factor(strep_tb$rad_numLabels,levels=rad_numLabels)
table(strep_tb$rad_numLabels)

#recode control as 0 and Streptomycin	 as 1
strep_tb$arm<-ifelse(strep_tb$arm=="Control", 0, 1)
strep_tb$arm<-factor(strep_tb$arm, levels=c(0:1), labels=c("Control", "Streptomycin"))
table(strep_tb$arm)
```

# 9. How many males and females are in the strep_tb dataset?
```{r}
#looking at males and females
table(strep_tb$gender) # There are 48 males and 59 females

```

# 10. Choose an appropriate model to model the association between arm (streptomycin vs. control) and improved (improvement vs. no improvement). a.Report the measure of association value and the 95% CI. b. Interpret the results.
```{r}
#Logistic regression
library(odds.n.ends)

improvedLogit <- glm(improved ~ arm, data=strep_tb, family="binomial")
summary(improvedLogit)

#calculate and print ORs and 95% CIs  
odds.n.ends(improvedLogit)

#Those taking streptomycin are 4.6 times the odds of controls of having their condition improved (95% CI 2.07, 10.63)



```

# 11. Choose an appropriate model to model the association between arm (streptomycin vs. control) and rad_num. a. Report the measure of association and value(s) and the 95% CIs. b. Interpret the results.
```{r}
library(nnet)

mod <- multinom(rad_numLabels ~ arm, data=strep_tb)
summary(mod)

#the odds of being in considerable deterioration vs dying increased by 1.25 in those taking streptomycin compared to controls
#the odds of being in moderate or slight deterioration vs dying increased by 0.38 in those taking streptomycin compared to controls
#the odds of having no material change vs dying increased by 0.84 in those taking streptomycin compared to controls
#the odds of having moderate or slight improvement vs dying increased by 0.99 in those taking streptomycin compared to controls
#the odds of having considerable vs dying increased by 3.19 in those taking streptomycin compared to controls


#calculate z score
z <- summary(mod)$coefficients/summary(mod)$standard.errors
# 2-tailed Z test
p <- (1 - pnorm(abs(z), 0, 1))*2
p

#Get confidence intervals for the log odds using confint function
ci <- confint(mod)
ci

exp(coef(mod))

#get confidence intervals for ORs
exp(confint(mod, level=0.95))

```

# Exercise 2.
This exercise will use the Framingham data to model the association between current smoking and death. The data dictionary can be found here: https://biolincc.nhlbi.nih.gov/media/teachingstudies/FHS_Teaching_Longitudinal_Data_Documentation_2021a.pdf?link_time=2021-12-02_12:02:38.620894. Please run the code below to get started.

```{r}
library(readr)
library(survminer) 
library(survival) 
library(ggfortify) 
library(lmtest)
#import the csv file
frm<-read_csv("frmgham2.csv")

frm$cursmoke<-factor(frm$CURSMOKE, levels= c(0:1), labels = c("Not Current Smoker", "Smoker"))
frm$sex<-factor(frm$SEX, levels=c(1:2), labels = c("Male", "Female"))
```

# 12. Determine number of individuals for variables needed for this analysis (cursmoke, sex, DEATH). Report the number of subjects who are a. current smokers, b. male, and c. who have died.
```{r}

table(frm$sex) #5022 males
table(frm$cursmoke) #6598 current smokers
table(frm$DEATH) #3527 have died

```

# 13. Determine the mean age and TIMEDTH. Report these means (a. age, b. TIMEDTH).
```{r}

mean(frm$TIMEDTH) #the mean days between baseline exam and death/censorship is 7,854
mean(frm$AGE) #the mean age is 54.8 years

```

# 14. Generate a Kaplan Meier plot that shows survival probabilities for death by cursmoke. Upload a snapshot or picture of your plot.
```{r}
install.packages("survival") 
install.packages('ggfortify') 
install.packages("survminer") 
library(survminer) 
library(survival) 
library(ggfortify)


cursmoke.surv <- survfit(Surv(TIMEDTH, DEATH) ~ cursmoke, frm) 
summary(cursmoke.surv) 

ggsurvplot(cursmoke.surv, data = frm, risk.table = TRUE, conf.int=TRUE, tables.theme = clean_theme())

```

# 15. Test for statistical differences between survival curves using the log rank test. Report and interpret the results.
```{r}

survdiff(Surv(TIMEDTH, DEATH) ~ cursmoke, data=frm)
#with a Chi-sq of 10 and 1 degree of freedom, there is a statistically significant different between survival curves of those who currently smoke and those who do not
```

# 16. Run a univariate Cox PH regression model to determine the hazard rate of DEATH in current smokers vs. non-current smokers. a. Report the HR and 95% CI and b. Interpret the results.
```{r}
pacman::p_load(survminer,survival,tidyverse,lmtest,foreign)

HRdeath.mod<-coxph(Surv(TIMEDTH, DEATH)~cursmoke, frm, ties="efron") #using ties = Efron, default is Efron, which is fine but this is how it would be changed.

summary(HRdeath.mod)

#Those who are current smokers have a 1.11 (95% CI 1.04-1.19) times higher hazard of death than those who are not current smokers.

```


# 17. Run a multivariable Cox model to determine the hazard rate of DEATH in current smokers vs. non-current smokers adjusted for age. a. Report the HR and 95% CI and b. interpret the results.
```{r}

HRdeathMV.mod<-coxph(Surv(TIMEDTH, DEATH)~cursmoke +AGE, frm)

summary(HRdeathMV.mod)

#After adjusting for age, hose who are current smokers have a 1.58 (95% CI 1.47-1.69) times higher hazard of death than those who are not current smokers.


```

# 18. Determine if sex modifies the association between the age-adjusted hazard rate of DEATH in current smokers vs. non-current smokers. Report the results of your statistical test and interpret the results.
```{r}

HRdeathINT.mod<-coxph(Surv(TIMEDTH, DEATH)~cursmoke + AGE + cursmoke*sex, frm)

summary(HRdeathINT.mod)

#There is no significant effect modification by (p for interaction=0.483) of the age-adjusted HR for the association between current smoking status and death by sex.

```

# 19. If sex modifies the association, run age-adjusted models to report associations stratified by sex. a. Report the age-adjusted HRs and 95% CIs for each sex and b. Interpret the results.
```{r}

#sex does not modify the association

```
# 20. Should the models be adjusted for sex or should sex-specific estimates be provided separately? Explain the rationale for your answer.

# Of note: the linearity assumption should be evaluated by age. However for the purpose of time, we have skipped that step in this exercise as well as other assumption tests.

# 21. Extra credit. Using the Cox model with the following right hand side: cursmoke + AGE + sex+ sex*cursmoke, determine the hazard rate of death in a 70 year old man who is a current smoker compared to a 80 year old man who is not a current smoker.  a. Report the HR and 95% CI. b. Interpret the results. c. How does this HR compare to the HR if they were both non-smokers?
```{r}

HRdeathADJ.mod<-coxph(Surv(TIMEDTH, DEATH)~cursmoke + AGE + sex+ sex*cursmoke, frm)

summary(HRdeathADJ.mod)

HR = cursmoke + age + sex + sex*cursmoke

#70 year old man, current smoker
= .39(1) + .07(70) + -.52(0)
=  .39 + 4.9  
exp(5.29) = 198.3434
# A 70 year old man who is current smoker has a hazard rate of death of 198.3434

#80 year old man, not current smoker
=.39(0) + .07(80) + -.52(0)
exp(80*0.07) = 270.4264

270.4264/198.3434
#The 80 year male non-smoker has 1.3 times the hazard rate of death compared to a 70 year old smoker

#Both smokers
#80 year old male smoker
= .39(1) + .07(80) 
exp(5.99) = 399.4146

399.4146/198.3434

#The 80 year male smoker has 2 times the hazard rate of death compared to a 70 year old smoker





```




